```{r echo=FALSE}
source("../R/create-Rmd-table.R")
```

```{r read metadata files, echo=FALSE}
file_metadata <- read.csv("../inst/metadata/file-metadata.csv", fileEncoding = "UTF-8-BOM")
column_metadata <- read.csv("../inst/metadata/column-metadata.csv", fileEncoding = "UTF-8-BOM")
column_category_metadata <- read.csv("../inst/metadata/column-category.csv", fileEncoding = "UTF-8-BOM")
```

# Model Parameter files

This section will go over the different model parameter files.

## Model Exports File

One of the four mandatory files that the user has to provide. Each row in this CSV file gives the location of a file needed for implementation. The metadata for this file is given below:

```{r echo=FALSE}
create_Rmd_table(
  "model-export",
  file_metadata,
  column_metadata,
  column_category_metadata
)
```
        
An example is provided below

```{r echo=FALSE}
model_export_file <- data.frame(
  fileType = c("variables", "variable-details", "model-steps"),
  filePath = c("./variables.csv", "../../variable-details.csv", "./model-steps.csv")
)

DT::datatable(
  model_export_file, 
  options = list(columnDefs = list(list(className = "dt-left", targets = "_all")))
)
```

Going into more details,

* The first row gives the location of the variables file. It's located in the same folder as the model export file and is called `variables.csv`
* The second row gives the location of the variable details file. It's located two folders above the model export file and is called `variable-details.csv`
* The third row gives the location of the model steps file. It's located in the same folder as the model export file and is called `model-steps.csv`

## Model Steps File

One of three mandatory files for this function, the other two are the variable details file and the variables file. This file specifies the steps for calculating the outcome of the model as well as their order. It does not specify the data for each step which will be defined in other files below. The metadata for this file is:

```{r echo=FALSE}
create_Rmd_table(
  "model-steps",
  file_metadata,
  column_metadata,
  column_category_metadata
)
```
    
An example is given below

```{r echo=FALSE}
eg_model_steps <- data.frame(
  "step" = c("dummy", "center"),
  "fileType" = c("N/A", "N/A"),
  "filePath" = c("dummy-data.csv", "../center-data.csv")
)

DT::datatable(eg_model_steps, options = list(columnDefs = list(list(className = "dt-left", targets = "_all"))))
```
    
The above example model steps file has two steps, dummy and center, defined in the order in which they should be done. 

The file which describes the dummy step is called `dummy-data.csv` and is in the same folder as the model steps file. 

The file which describes the center step is called `center-data.csv` and is in the folder above the model steps file.

Notice how the `fileType` columns for both these steps is `N/A` or `Not Applicable`. Since the dummy and center steps require only one file to describe them, we do not need to provide a value for the `fileType` columns for those rows. The fine and gray model step described in this document required multiple files to describe it, hence requiring the use of the `fileType` column.

## Variable Details File

One of the mandatory files. Its structure is defined here https://big-life-lab.github.io/cchsflow/articles/variable_details.html.

## Variables File

One of the mandatory files that defines the starting variables in the model. Its columns have been defined here https://big-life-lab.github.io/cchsflow/articles/variables_sheet.html.

## Dummy Step File

Dummying is a transformation where a categorical variable with greater than n > 2 categories, is converted to a set of n dicotomous variables. Information about how this transformation is done can be seen [here](https://en.wikiversity.org/wiki/Dummy_variable_(statistics)).

The metadata for this file is provided below:

```{r echo=FALSE}
create_Rmd_table(
  "dummy",
  file_metadata,
  column_metadata,
  column_category_metadata
)
```
    
An example is given below

```{r echo=FALSE}
eg_dummy_step <- data.frame(
  "origVariable" = c("smoker_type", "smoker_type", "smoker_type", "drinker_type", "drinker_type", "drinker_type", "drinker_type"),
  "catValue" = c(0,1,2, 0,1,2,3),
  "dummyVariable" = c("non_smoker", "current_smoker", "former_smoker", "non_drinker", "current_light_drinker", "current_heavy_drinker", "former_drinker")
)

DT::datatable(eg_dummy_step, options = list(columnDefs = list(list(className = "dt-left", targets = "_all"))))
```

The example above describes the data for dummying two variables:

* smoker_type
* drinker_type

The `smoker_type` variable has three dummy variables:

* `non_smoker`: Represents category 0 in the `smoker_type` variable
* `current_smoker`: Represents category 1 in `smoker_type` variable
* `former_smoker`: Represents category 2 in the `smoker_type` variable

The `drinker_type` variable has four dummy variables:

* `non_drinker`: Represents category 0 in the `drinker_type` variable
* `current_light_drinker`: Represents category 1 in the `drinker_type` variable
* `current_heavy_drinker`: Represents category 2 in the `drinker_type` variable
* `former_drinker`: Represents category 3 in the `drinker_type` variable

## Centering Step File

Centering is a transformation where the original variable is subtracted by a value, usually it's mean or median. The metadata for this file is given below:

```{r echo=FALSE}
create_Rmd_table(
  "center",
  file_metadata,
  column_metadata,
  column_category_metadata
)
```
    
An example is given below,

```{r echo=FALSE}
eg_centering_step <- data.frame(
  "origVariable" = c("age", "diet_score"),
  "centerValue" = c(40, 4),
  "centeredVariable" = c("age_C", "diet_score_C"),
  "centeredVariableType" = c("cont", "cont")
)

DT::datatable(eg_centering_step, options = list(columnDefs = list(list(className = "dt-left", targets = "_all"))))
```

The example above describes the data for centering two variables, `age` and `diet_score`

The values used to center the two variables are `40` and `4` respectively

The names of the new centered variables are `age_C` and `diet_score_C` and they are both continuous

## Restricted Cubic Spline (RCS) Step File

The metadata for this file is given below:

```{r echo=FALSE}
create_Rmd_table(
  "rcs",
  file_metadata,
  column_metadata,
  column_category_metadata
)
```
    
An example is given below,

```{r echo=FALSE}
eg_rcs_step <- data.frame(
  "variable" = c("Age_c", "PackYears_c"),
  "rcsVariables" = c("AgeC_rcs1;AgeC_rcs2;AgeC_rcs3;AgeC_rcs4", "PackYearsC_rcs1;PackYearsC_rcs2"),
  "knots" = c("-11.5;-6.5;-1.5;5.5;16.5", "-21.2;-10.35;34.8")
)

DT::datatable(eg_rcs_step, options = list(columnDefs = list(list(className = "dt-left", targets = "_all"))))
```

The example describes the data for creating new spline variable for two variables, `Age_c` and `PackYears_c`

For `Age_c`, a 5 knot spline is used and 4 new spline variables are created. The knots given are,

1. -11.5
2. -6.5
3. -1.5
4. 5.5
5. 16.5

The new variables created are,

1. AgeC_rcs1
2. AgeC_rcs2
3. AgeC_rcs3
4. AgeC_rcs4

Notice how in the CSV, each knot is seperated by a ";" and each new variable name is also seperated by a ";"

Similarly, for `PackYears_c`, we define a 3 knot spline where the knots are `-21.2`, `-10.35` and `34.8`. The new spline variables are `PackYearsC_rcs1` and `PackYearsC_rcs2`.

## Interaction Step File

Theoretically, an interaction variable is created to model effect modification. Effect modification occurs when the magnitude of the effect of a risk factor on a outcome differs depending on another risk factor. An example is age, cancer and their effect on mortality, where as you get older the effect that cancer has on your risk of mortality increases. Here, age is said to modify the effect that cancer has on mortality. Mathematically, an effect modification is modeled by creating a new variable whose value is the product of the two "interacting" variables, e.g., age and cancer.

The metadata for this file is given below:

```{r echo=FALSE}
create_Rmd_table(
  "interaction",
  file_metadata,
  column_metadata,
  column_category_metadata
)
```
    
An example is given below,

```{r echo=FALSE}
eg_interaction_step <- data.frame(
  "interactionVariable" = c("AgeXCancer", "AgeXHypertension"),
  "interactingVariables" = c("Age;Cancer", "Age;Hypertension"),
  "interactionVariableType" = c("cont", "cont")
)

DT::datatable(eg_interaction_step, options = list(columnDefs = list(list(className = "dt-left", targets = "_all"))))
```

The example above creates two new interaction variables, `AgeXCancer` and `AgeXHypertension`. `AgeXCancer` is created using the `Age` and `Cancer` variables while `AgeXHypertension` is created using the `Age` and `Hypertension` variables. Both new interaction variables are continuous. Notice how the interacting variables are seperated by a semi-colon in the example.

## Fine and Gray Model Step Files

Similar to a cox regression model, a fine and gray model estimates the risk of an event occuring at some time in the future, the difference being that it takes competing risks into account. Specifying this step requires two files,

1. A beta coefficients file which specifies the names of the covariates in the model and their beta coefficient
2. A baseline hazards file which specifies a time value and the baseline hazard value associated with it

The metadata for the beta coefficients file is given below:

```{r echo=FALSE}
create_Rmd_table(
  "beta-coefficients",
  file_metadata,
  column_metadata,
  column_category_metadata
)
```
    
The metadata for the baseline hazards file is given below:

```{r echo=FALSE}
create_Rmd_table(
  "baseline-hazards",
  file_metadata,
  column_metadata,
  column_category_metadata
)
```

The baseline hazards file has the following columns:

1. **time**: The time upto which this baseline hazard should be used
    **type**: number
2. **baselineHazard**: The baseline hazard value for this time
    **type**: number
    
When adding a fine and gray model step, make sure to add a variable in the variable details sheet called time. This variable represents the time interval when the model is valid. It should be a continuous variable whose `recTo` values represents the time interval. It's `units` column should specify the time metric being used for example `years`. For example, a model that can predict risk at any point in time between 1 year and 5 years from today would have the following entry in the variable details sheet,

```{r example time variable, echo=FALSE}
eg_time_variable <- data.frame(
  "variable" = c("time", "time"),
  "recTo" = c(1, 5),
  "catLabel" = c("Start Time", "End Time"),
  "catLabelLong" = c("Start Time", "End Time"),
  "units" = c("years", "years")
)

DT::datatable(eg_time_variable, options = list(columnDefs = list(list(className = "dt-left", targets = "_all"))))
```

The values of the other columns in the variable details sheet for this variable should be `N/A`.

An example model steps file and the referenced fine and gray model files are shown below.

```{r example fine and gray model, echo=FALSE}
eg_fine_and_gray_model_steps_file <- data.frame(
  "step" = c("fine-and-gray", "fine-and-gray"),
  "fileType" = c("beta-coefficients", "baseline-hazards"),
  "filePath" = c("beta-coefficients.csv", "baseline-hazards.csv")
)
DT::datatable(eg_fine_and_gray_model_steps_file, options = list(columnDefs = list(list(className = "dt-left", targets = "_all"))))

eg_beta_coefficients_csv <- data.frame(
  "variable" = c("Age", "Sex"),
  "coefficient" = c(0.01, 2)
)
DT::datatable(eg_beta_coefficients_csv, options = list(columnDefs = list(list(className = "dt-left", targets = "_all"))))

eg_baseline_hazards_csv <- data.frame(
  "time" = c(1, 2, 3, 4, 5),
  "baselineHazard" = c(0.1, 0.2, 0.3, 0.4, 0.5)
)
DT::datatable(eg_baseline_hazards_csv, options = list(columnDefs = list(list(className = "dt-left", targets = "_all"))))
```

The model steps file has 2 rows, both of which are a fine and gray step. The type of file described in each row is is given by `fileType` column. The first row references the a beta coefficients file and the second row references a baseline hazards file. The path to each file is given in the `filePath` column and once again should be relative to the model steps file that described them.

The beta coefficients file says that the model has two covariates, `Age` and `Sex` and specifies the beta coefficients for each one which are `0.01` and `2` respectively.

The baseline hazards has 5 rows and describes the baseline hazards for each year in the model. Notice how the lowest time value is `1` and the highest value is `5`. This is because of the time variable we specified earlier which goes from 1 to 5.